---
title: "Getting Started with the fcds Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fcds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[fcds-miami]: https://fcds.med.miami.edu/
[fcds-request]: https://fcds.med.miami.edu/inc/datarequest.shtml

## Installation

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("GerkeLab/fcds")
```

## Setup

```{r setup, message=FALSE}
library(dplyr)
library(fcds)
```

The **fcds** package doesn't include data from the [FCDS Florida Cancer Registry][fcds-miami].
Please visit the [FCDS Data Request][fcds-request] page to request the most recent Florida Statewide Cancer Registry data.

Once you have downloaded the data, you can use **fcds** to pre-process and import the FCDS data into R.

```{r eval=FALSE}
fcds <- fcds_import("STAT_dataset_2018.dat")
```

The imported, pre-processed data is automatically cached in the `.fcds` directory in your home directory for later use.
Calling `fcds_load()` will re-load the cached data.

```{r eval=FALSE}
fcds <- fcds_load()
```

You can customize the folder where the cached data files are stored; see `?fcds_default_data_path` for more information.

You can also list the files currently stored in the data cache with `fcds_cache_ls()` and you can clean outdated files with `fcds_cache_clean()`, which by default retains the most recently imported FCDS data file.

### Example Data

If you don't have access to the FCDS data, we have provided a small example data set with similar properties to the FCDS data, but without any overlap in the attributes of cases.

```{r}
fcds <- fcds::fcds_example

fcds[1:5, ]
```

## Looking up FCDS Values

### FCDS Constants

Lookup the values of variables in the imported data (referred to as a `labels` of the original data `value`) with `fcds_const()`

```{r}
fcds_const(NULL)

fcds_const("cancer_status")
```

If you supply `full = TRUE`, the original column name (`name_original`) and `value` from the FCDS data will be reported alongsize the cleaned column name (`name_clean`) and the value `label` used by the **fcds** package.

```{r}
fcds_const("cancer_site_group", full = TRUE)
```

### FCDS Variables

All of the variables in the imported FCDS data are documented in `?fcds_import`.
You can also list common groups of `fcds` variables using `fcds_vars()`.

```{r}
fcds_vars("demo")

fcds_vars("cancer")
```

These column names can be used in conjunction with `dplyr::select()`, or you can supply FCDS data to `fcds_vars()` using the `.data` argument to select matching columns.

```{r}
fcds %>% select(fcds_vars("id", "demo"))

fcds_vars(.data = fcds, "id", "icdo3")
```

## A Worked Example: Prostate Cancer

**Note:** This data uses the `fcds_example` data and therefore bears no resemblance to reality.
The examples herein are only relevant to the demonstration of the functions of the **fcds** package.

```{r}
fcds_prostate <- 
  fcds %>% 
  # Filter to Prostate Cancer in two Florida counties
  filter(
    cancer_site_group == "Prostate Gland",
    county_name %in% c("Hillsborough", "Pinellas"),
    sex == "Male"
  ) %>% 
  # Count incidences of Prostate Cancer
  count_fcds(county_name, cancer_site_group) %>% 
  # Fill in missing age groups w.r.t. specified columns
  complete_age_groups(
    county_name, cancer_site_group, tidyr::nesting(year_group, year)
  ) %>% 
  # Calculate age-adjusted rate from incidence
  age_adjust() %>% 
  # Compute average yearly age-adjusted rate over 5 year range
  mutate(rate = rate / 5)

fcds_prostate
```

This data can readily be passed to `ggplot2` for visualization.

```{r fig.height=3, fig.width=7}
library(ggplot2)

ggplot(fcds_prostate) +
  aes(year_group, rate, color = county_name, group = county_name) +
  geom_line() +
  geom_point()
```
