---
title: "FCDS Age Adjustment Notes"
date: '`r strftime(Sys.time(), "%A, %b %d, %Y")`'
mainfont: 'Source Serif Pro'
monofont: 'Source Code Pro'
output:
  github_document: default
  rmarkdown::html_vignette:
    df_print: kable
    standalone: true
  pdf_document: 
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, cache = FALSE,
  warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7, 
  fig.showtext = TRUE # for the fancy fonts, disable if not needed
)
```

```{css global-css, echo=FALSE}
img {
  margin: 0;
  padding: 0;
  max-width: 100%;
}
```

```{r library, eval=FALSE}
library(tidyverse)

browser()
# Use the grkmisc theme
theme_set(
  grkmisc::theme_grk(
    base_family = "Fira Sans",
    axis_text_family = "Fira Sans Condensed",
    axis_title_family = "Fira Sans Condensed",
    default_geom_font = "Fira Sans Condensed",
    use_showtext = TRUE
  ) + 
    theme(panel.grid.minor = element_blank())
)
```

```{r load, include=FALSE}
# Load Data
```

<!-- Start Document Here -->

## U.S. Cancer Statistics Data Visualizations Tool Technical Notes

<https://www.cdc.gov/cancer/uscs/pdf/uscs-datavisualizationtool-technicalnotes-508.pdf>

## County-Level Population Files

<https://seer.cancer.gov/popdata/download.html>

> The U.S. Census Bureau annually releases unbridged population estimates for five-year age groups and race at the county level. The Census Bureau does not release bridged race estimates by single year of age at the county level due to concerns about the reliability of these estimates. However, these estimates are provided to the National Cancer Institute to meet programmatic needs such as the creation of age groupings that differ from the standard groupings used by the Census Bureau. Users of the single-year-of-age county-level bridged race population estimates should carefully consider the limited reliability of these estimates.

## Age Adjustment

An age-adjusted rate is a weighted average of the age-specific (crude) rates according to the proportion of the population in the corresponding age group of a standard population.

The age adjusted rates are calculated according to

$$\left. r_{adj} \right\vert_x^y = \sum_{\text{age}=x}^y \left\lbrack \frac{count_{\text{age}}}{pop_{\text{age}}} \times \frac{stdpop_{\text{age}}}{\sum_{j=x}^{y} stdpop_j} \right\rbrack$$

This was implemented in code as:

```{r eval=FALSE, echo=TRUE}
age_adjust <- function(
  .data,
  outcome_var = n,
  year_var = dx_year_mid,
  fl_pop = get_data("seer_fl_pop"),
  std_pop_data = get_data("seer_std_ages"),
  keep_age_group = FALSE
) {
  outcome_var <- rlang::enquo(outcome_var)
  outcome_var_name <- rlang::quo_name(outcome_var)
  year_var <- rlang::enquo(year_var)
  year_var_name <- rlang::quo_name(year_var)

  # Age and Year are required in source data
  stopifnot(year_var_name %in% names(.data))
  stopifnot("age_group" %in% names(.data))
  .data <- filter(.data, age_group != "Unknown")

  #  Get groups from .data
  #  Roll up FL and std pop data for those groups
  #  Merge into fcds
  #  Calculate age-adjusted rate
  .data <- merge_fl_pop(.data, !!year_var, fl_pop)

  data_groups <- setdiff(group_vars(.data), year_var_name)
  .data <- .data %>%
    group_by(!!!rlang::syms(data_groups))

  if (keep_age_group) return(.data)
  .data %>% age_adjust_finalize(!!outcome_var, std_pop_data)
}

merge_fl_pop <- function(
  .data,
  year_var = dx_year_mid,
  fl_pop = get_data("seer_fl_pop")
) {
  year_var <- rlang::enquo(year_var)
  year_var_name <- rlang::quo_name(year_var)
  stopifnot(year_var_name %in% names(.data))

  data_groups <- setdiff(group_vars(.data), year_var_name)
  fl_groups <- c(intersect(data_groups, names(fl_pop)), "year")
  if ("age_group" %in% fl_groups) {
    fl_groups <- rev(union("age_group", rev(fl_groups)))
  }

  fl_pop <-
    fl_pop %>%
    group_by(., !!!rlang::syms(fl_groups)) %>%
    summarize(population = sum(population)) %>%
    rename(!!year_var_name := year)

  left_join(.data, fl_pop, by = common_names(.data, fl_pop))
}

age_adjust_finalize <- function(
  .data,
  outcome_var = n,
  std_pop_data = get_data("seer_std_ages")
) {
  outcome_var <- rlang::enquo(outcome_var)

  age_groups <- unique(.data$age_group)

  # Get standard population
  .data <-
    std_pop_data %>%
    filter(age_group %in% age_groups) %>%
    select(age_group, std_pop) %>%
    mutate(w = std_pop / sum(std_pop)) %>%
    left_join(.data, ., by = "age_group")

  # Get data groups and make sure "age_group" is last for summary
  data_groups <- group_vars(.data)
  data_groups <- setdiff(data_groups, "age_group")

  .data %>%
    ungroup() %>%
    mutate(
      rate = !!outcome_var / population * w
    ) %>%
    group_by(!!!rlang::syms(data_groups)) %>%
    summarize_at(vars(n, population, rate), sum) %>%
    group_by(!!rlang::sym(data_groups[length(data_groups)]), add = TRUE) %>%
    mutate(rate = rate * 100000)
}
```
